---
title: "Quickstart: building a synthetic control with `Synth`"
author: "Jens Hainmueller and Alexis Diamond"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Quickstart: building a synthetic control with Synth}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 96
)
```

This vignette is the five-minute version: how to build, fit, and inspect
a synthetic control with `Synth`. For inference (prediction intervals,
placebo p-values), see `vignette("inference", package = "Synth")`.

## 1. Build the inputs

The recommended path for new users is `synth_data()` — a one-line
wrapper around `dataprep()` that picks sensible defaults. We use the
canonical Basque-country example (Abadie & Gardeazabal 2003).

```{r setup}
library(Synth)
data(basque)
```

```{r dataprep}
dp <- synth_data(
  panel              = basque,
  outcome            = "gdpcap",
  unit_col           = "regionno",
  time_col           = "year",
  treated            = 17,                              # Basque country
  controls           = c(2:16, 18),                     # other regions
  treatment_time     = 1970,
  predictors         = c("school.illit", "school.prim",
                         "invest"),
  special_predictors = list(
    list("gdpcap",          1960:1969,           "mean"),
    list("sec.agriculture", seq(1961, 1969, 2),  "mean")
  ),
  unit_names_col     = "regionname"
)
```

`synth_data()` returns a `dataprep`-shaped list, so anything downstream
(`synth()`, `path.plot()`, `synth_inference()`, etc.) works exactly
like the long-form `dataprep()` output.

If you need full control over the column-by-column construction,
`?dataprep` is still there.

## 2. Fit the synthetic control

```{r synth}
fit <- synth(dp, verbose = FALSE)
```

`fit$solution.w` is the donor weights (sum to 1, all in `[0, 1]`).
`fit$solution.v` is the predictor weights chosen by the V-search.

## 3. Inspect

`synth.tab()` produces a balance table comparing the treated unit to
its synthetic control on each predictor:

```{r tabs}
tabs <- synth.tab(synth.res = fit, dataprep.res = dp)
tabs$tab.pred
```

`path.plot()` shows the treated unit and its synthetic control over
time:

```{r path}
path.plot(synth.res = fit, dataprep.res = dp,
          Ylab = "Real per-capita GDP",
          Xlab = "Year",
          Legend = c("Basque country", "Synthetic Basque country"),
          tr.intake = 1970)
```

`gaps.plot()` shows the gap (treated minus synthetic):

```{r gaps}
gaps.plot(synth.res = fit, dataprep.res = dp,
          Ylab = "Gap in GDP per capita",
          Xlab = "Year",
          tr.intake = 1970)
```

## 4. What next?

- For prediction intervals around the synthetic counterfactual, see
  `?synth_inference` and `vignette("inference", package = "Synth")`.
- For placebo-based inference (Abadie–Diamond–Hainmueller 2010), see
  `?generate_placebos` and `?mspe_test`.
- For ggplot2-friendly plots, `library(ggplot2)` and call
  `autoplot()` on the inference / placebo objects.
- For the long-form constructor, see `?dataprep`.
